FISTA with adaptive discretisations

### 2021-01-21 paper.tex
slightly modified the introduction

\section{Introduction}
The Fast Iterative Shrinkage-Thresholding Algorithm (FISTA) was first proposed by \citet{Beck2009} and is now a very popular algorithm for minimising the sum of two convex functions. We will write this as the problem of computing 
\begin{equation}\label{eq: min F = fD+fC}
	u^* \in \argmin_{\varf0\in\F H}\Func0(\varf0)  \qquad\text{such that}\qquad \Func0(\varf0)\coloneqq \func1(\varf0) + \func2(\varf0),
\end{equation}
where $\func1\colon\F H\to\R$ is a convex differentiable function with $L$-Lipschitz gradient and $\func2\colon\F H\to\bar\R$ is a convex function. The iterates of the FISTA algorithm will be denoted $\varf0_n\in\F H$. \citet{Beck2009} showed that $\Func0(\varf0_n) - \Func0(\varf0^*)$ converges at the optimum rate of $n^{-2}$ when $u^*\in \F H$ is finite dimensional and \cite{Chambolle2015} extended this result to the general Hilbert space setting. Many further works have gone on to demonstrate even faster practical convergence rates for slightly modified variants of FISTA \citep{Tao2016,Liang2017,Alamo2019}.

@@

\subsection{Canonical example}
While the analysis of this work is applicable to general choices of $\func1$ and $\func2$, we will refer back to the example of the (Beurling) Lasso model both for motivation and for numerical examples. We state the Lasso energy as
\begin{equation}\label{eq: Lasso energy intro}
	\Func0(\varf0) = \tfrac12\norm{\A\varf0-\vdata0}_2^2 + \mu\norm{\varf0}_1,
\end{equation}
where $\vdata0\in\R^m$ is some observed data, $\A\colon \C M(\Domain)\to \R^m$ is the forward map and $\mu>0$ is an appropriately chosen scalar. We have $\F H = L^2(\Domain)$ and $\spcf0 = \C M(\Domain)$ the space of measures on $\Domain$ equipped with the $L^1$ norm. In the continuous setting $\Domain = [0,1]^d$ and for discrete we have $\Domain=\F N$, i.e. $\spcf0\cong\ell^1(\R)$.

It is a well-known result of representer theorems for $\Func0$ \citep{Unser2016,Boyer2019} that there exist minimisers of the form 
\begin{equation}
	\varf0^* = \sum_{i=1}^{m}\vars0_i\delta_{\vvarx0_i}, 
\end{equation}
for some $\vars0_i\in\R$ and where $\delta_{\vvarx0_i}$ is the Dirac delta function centred at $\vvarx0_i\in\R^d$.

This is a practical example where we have the relation $\varf0^*$ does not lie in the Hilbert space $\F H$, and yet it is very easy to approximate with a discrete basis. Our approach will be to use finite elements which can clearly generalise to other examples.

@@

\section{Lasso minimisation}\label{sec: Lasso definition}
We now return to the concrete example of Lasso which will be used for numerical results in \Cref{sec: numerics}. We consider three forms of Lasso which will be referred to as continuous, countable, and discrete Lasso depending on whether the space $\spcf0$ is $\C M([0,1]^d)$, $\ell^1(\R)$, or finite dimensional respectively. In each case, the energy can be written as 
\begin{equation}\label{eq: Lasso energy}
	\Func0(\varf0) = \tfrac12\norm{\A\varf0-\vdata0}_{\ell^2}^2 + \mu\Norm{\varf0}
\end{equation}
for some $\A\colon \spcf0\cup\F H\to\R^m$ and $\mu>0$, where $\Norm\cdot=\norm\cdot_1$.

@@

\begin{figure}[H]\centering
	\includegraphics[width=.84\textwidth]{lasso_reduced_convergence}
	\caption{Continuous convergence of adaptive (coarse-to-fine pixel size) compared with uniform discretisation (constant pixel size) with respect to number of iterations. }\label{fig: comparison with iteration}

@@

	\includegraphics[width=.84\textwidth]{lasso_convergence_time}
	\caption{Continuous convergence of adaptive compared with uniform discretisation with respect to wall-clock time and total number of pixels (memory requirement).}\label{fig: comparison with time}
\end{figure}

\paragraph{Tree structure of wavelets}
Finite elements are constructed with a mesh which provided a useful tool for adaptive refinement in \Cref{sec: bound continuous}. For wavelets, we will associate a tree with every discretisation and the leaves of the tree form a mesh. This perspective comes from the multi-resolution interpretation of wavelets. An example is seen in \Cref{fig: wavelet tree} for 1D Haar wavelets, $w_{j,k}(\varx0) = \sqrt{2}^k\vard1(2^{k}\varx0-j)$ where $\vard1 = \1_{[0,1)} - \1_{[-1,0)}$. 
\begin{figure}\begin{center}\begin{tikzpicture}[grow'=right]
			\tikzstyle{level 1}=[level distance=5em, sibling distance=0em];
			\tikzstyle{level 2}=[level distance=5em, sibling distance=6em];
			\tikzstyle{level 3}=[level distance=5em, sibling distance=3em];
			\tikzstyle{bag} = [text width=3em, text centered];

@@

	\end{tikzpicture}\end{center}
	\caption{Tree representation of 1D wavelets $w_{j,k}$ are arranged in a tree structure with their support underneath.}\label{fig: wavelet tree}
\end{figure}
In higher dimensions, the only two things which change are the number of children ($2^d$ for non-leaves) and at each node you store the coefficients of $2^d-1$ wavelets. The support on each node is still a disjoint partition of unity consisting of regular cubes of side length $2^{-k}$ at level $k$. The only change in our own implementation is to translate the support to $[-\tfrac12,\tfrac12]^2$. We briefly remark that the tree structuring of wavelets is not novel and appears more frequently in the Bayesian inverse problems literature, for example see \citet{Castillo2019}.

@@

\begin{figure}[H]\centering
	\includegraphics[width=.85\textwidth]{haar_data}
	\caption{Phantoms and data used for wavelet-sparse tomography optimisation. The Shepp-Logan data is exact but the data for the disc-phantom has \SI{5}{\percent}\ Gaussian white noise. Without noise the data would be uniform with respect to the angle.}\label{fig: haar data}

@@

\end{document}

>>>>

\section{Introduction}
The Fast Iterative Shrinkage-Thresholding Algorithm (FISTA) was proposed by \citet{Beck2009}
as an extension of Nesterov's fast gradient method (see~\cite{Nesterov2004})
and is now a very popular algorithm for minimising the sum of two convex functions.
We write this as the problem of computing 
\begin{equation}\label{eq: min F = fD+fC}
	u^* \in \argmin_{\varf0\in\F H}\Func0(\varf0)  \qquad\text{such that}\qquad \Func0(\varf0)\coloneqq \func1(\varf0) + \func2(\varf0),
\end{equation}
where $\func1\colon\F H\to\R$ is a convex differentiable function with $L$-Lipschitz gradient and $\func2\colon\F H\to\bar\R$ is a convex function, whose ``proximity operator'' is easy to compute.
The iterates of the FISTA algorithm will be denoted $\varf0_n\in\F H$. \citet{Beck2009} showed that $\Func0(\varf0_n) - \Func0(\varf0^*)$ converges at the optimum rate of $n^{-2}$.
% when $u^*\in \F H$ is finite dimensional and
\cite{Chambolle2015} showed (up to a minor modification) convergence of the iterates in
a general Hibert space setting.
% extended this result to the general Hilbert space setting.
Many further works have gone on to demonstrate %even
faster practical convergence rates for slightly modified variants of FISTA \citep{Tao2016,Liang2017,Alamo2019}.

@@

@@

\section{Lasso minimisation}\label{sec: Lasso definition}
We now focus on the concrete example of Lasso which will be used for numerical results in \Cref{sec: numerics}. We consider three forms of Lasso which will be referred to as continuous, countable, and discrete Lasso depending on whether the space $\spcf0$ is $\C M([0,1]^d)$, $\ell^1(\R)$, or finite dimensional respectively. In each case, the energy can be written as 
\begin{equation}\label{eq: Lasso energy}
	\Func0(\varf0) = \tfrac12\norm{\A\varf0-\vdata0}_{\ell^2}^2 + \mu\Norm{\varf0}
\end{equation}
for some $\A\colon \spcf0\cup\F H\to\R^m$ and $\mu>0$, where $\Norm\cdot=\norm\cdot_1$.

@@

\begin{figure}\centering
	\includegraphics[width=.84\textwidth]{lasso_reduced_convergence}
	\caption{Continuous convergence of adaptive (coarse-to-fine pixel size) compared with uniform discretisation (constant pixel size) with respect to number of iterations. }\label{fig: comparison with iteration}

@@

	\includegraphics[width=.84\textwidth]{lasso_convergence_time}
	\caption{Continuous convergence of adaptive compared with uniform discretisation with respect to wall-clock time and total number of pixels (memory requirement).}\label{fig: comparison with time}

	\begin{center}\begin{tikzpicture}[grow'=right]
		\tikzstyle{level 1}=[level distance=5em, sibling distance=0em];
		\tikzstyle{level 2}=[level distance=5em, sibling distance=6em];
		\tikzstyle{level 3}=[level distance=5em, sibling distance=3em];
		\tikzstyle{bag} = [text width=3em, text centered];

@@

	\end{tikzpicture}\end{center}
	\caption{Tree representation of 1D wavelets $w_{j,k}$ are arranged in a tree structure with their support underneath.}\label{fig: wavelet tree}
\end{figure}

\paragraph{Tree structure of wavelets}
Finite elements are constructed with a mesh which provided a useful tool for adaptive refinement in \Cref{sec: bound continuous}. For wavelets, we will associate a tree with every discretisation and the leaves of the tree form a mesh. This perspective comes from the multi-resolution interpretation of wavelets. An example is seen in \Cref{fig: wavelet tree} for 1D Haar wavelets, $w_{j,k}(\varx0) = \sqrt{2}^k\vard1(2^{k}\varx0-j)$ where $\vard1 = \1_{[0,1)} - \1_{[-1,0)}$.

In higher dimensions, the only two things which change are the number of children ($2^d$ for non-leaves) and at each node you store the coefficients of $2^d-1$ wavelets. The support on each node is still a disjoint partition of unity consisting of regular cubes of side length $2^{-k}$ at level $k$. The only change in our own implementation is to translate the support to $[-\tfrac12,\tfrac12]^2$. We briefly remark that the tree structuring of wavelets is not novel and appears more frequently in the Bayesian inverse problems literature, for example see \citet{Castillo2019}.

@@

\begin{figure}\centering
	\includegraphics[width=.85\textwidth]{haar_data}
	\caption{Phantoms and data used for wavelet-sparse tomography optimisation. The Shepp-Logan data is exact but the data for the disc-phantom has \SI{5}{\percent}\ Gaussian white noise. Without noise the data would be uniform with respect to the angle.}\label{fig: haar data}

@@

\end{document}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
